On the Characterizations of Rational Geodesics on the Infinite Necker Cube Surface

### 2018-01-04 finalpaper.tex
010318

\documentclass[]{article}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{svg}
\usepackage{enumitem}
\usepackage{color}
\usepackage{float}
\usepackage[utf8]{inputenc}
\usepackage{wrapfig}
\usepackage[arrow]{xy}
\usepackage{tikz-cd}

@@

\title{Periodicity of Geodesics on the Necker Cube Surface}
\author{Pavel Javornik}

\maketitle

%\begin{center}
%
%\includesvg[width=4.8in]{cubecoverphoto}\\
%\end{center}

\begin{abstract}
\noindent Although it is most recognizable as M.C. Escher's adaptation of an optical illusion popularized by Louis Albert Necker, the Necker cube surface has a complex structure and is, in analaytic language, described as an infinite, flat (locally isometric to $\RR^2$ on smooth sections), $\RR^3$-embeddable Euclidean cone surface with countably many singularities of cone angles $3\pi$ and $\frac{3\pi}{2}$. Its geometric construction is described as the gluing of infinitely many unit cubes along their edges so that every face shares an edge with exactly four others. A unit-speed geodesic on this surface has an initial trajectory angle $\theta\in\RR/2\pi\ZZ$ given up by the surface's local plane isometry. Given a rational vector direction in $\RR^2$ (components are relatively prime integers) obtained from $\theta$, one can classify every periodic/drift-periodic rational geodesic on the surface. In addition its total period/length and, in the case where it is drift-periodic, translational drift as Euclidean distance in its ambient space is given by solving a system of Diophantine equations seemingly related to Farey sequences.
\end{abstract}
\begin{figure}[H]
\centering
\includesvg[width=4in]{cubesxyzpaths2bwv2b}
\label{fig:front}
\caption{Periodic and drift-periodic geodesics on the Necker cube surface}
\end{figure}
\section{Introduction}
The Necker cube has made numerous appearances in the work of mathematicians, crystallographers, and scientists interested in human visual systems even before appearing in the works of M.C. Escher [cite]. Louis Albert Necker was credited for having discovered this optical illusion and studying its various properties [cite]. The solid presentation of the cube, when rendered as a flat surface with three rear faces removed, achieves a similar effect when its three visible faces are shaded in a particular way not unlike the illusion given off by the ``wire-frame" cube.[cite] The Necker cube surface is then obtained by gluing infinitely many copies of the semi-cube surface along its outer edges (Fig 1).

Dynamical classification of geodesics on flat, infinite surfaces admitting cone singularities such as the Necker cube surface ultimately boils down to studying the surface's and, if it exists, base surface's various symmetries. The rotational holonomy group acts on the unit tangent bundle of the surface as a unit vector (derivative of the geodesic at any given time) undergoes parallel transport along the geodesic path. If a surface is periodically constructed and obtained as a normal cover of some elementary base surface (like a polygon with edge identifications), this holonomy group might be given up by observing how the push-forward of elements in the group of deck transformations acts on tangent spaces. In particular, loops on polyhedral surfaces punctured at its  cone singularities (cone angles that are rational multiples of $\pi$) tend to have discrete holonomy groups as its relative holonomy (holonomy of contractible loops) is trivial.

\begin{rem}
Familiarity is assumed on the part of the reader with covering space theory, translation (particularly Veech) surfaces, and their associated Veech groups. For general surveys on these topics we refer readers to: [cite], [cite].
\end{rem}

\subsection{Discussion of Results}
Our initial experiments supported the theory that there would be a strong correlation between a choice of initial trajectory angle and the dynamical properties of a geodesic traveling on the surface. [throw in some trajectories in the appendix] A surface composed of infinitely many cubes is endowed with a flat metric, i.e. every neighborhood is locally isometric to $\RR^2$. Via the parallel transport of a unit vector tangent to the surface in the direction of the geodesic, we view its image as a sequence of straight lines on the (smooth) faces of the surface that tightly wrap around its sharp edges (Fig 1).
\\
\begin{wrapfigure}{l}{2in.}
\includesvg{vectorplane}
\end{wrapfigure}
\noindent Let $u_0$ be a point on the surface that is contained in the interior of a face and consider a tangent unit vector $v\in\RR^3$ based at point $u_0$. Each face is parallel to a $2$-dimensional subspace of $\RR^3$ spanned by exactly two basis vectors, so one component of $v$ is always 0. Thus $v$ is identified with $v_0\in\RR^2$.
\\\\\\
Denote the Necker cube surface $\bUs$. The angle $v_0$ makes relative to a choice of basis is the \emph{initial trajectory angle} of a \emph{unit-speed geodesic} based at $u_0$ on the surface, $\Phi_t:\RR\rightarrow\bUs$. The initial trajectory angle is said to be $\emph{rational}$ if there is some $k\in\RR$ such that $kv_0$'s nonzero components are relatively prime integers. A rational initial trajectory angle in $\RR^2$ falls into one of two categories:
\begin{Def}
Let $v_0$ be a rational unit vector of the form $\frac{1}{k}(x,y)\in\RR^2$ with $x,y\in\mathbb{Z}$ and $k=\sqrt{x^2+y^2}\in\mathbb{R}$. We say $kv_0$ is an \textbf{odd-odd} vector if $x,y$ are relatively prime and both odd. We denote the \textbf{set of all odd-odd directions} by $\mathcal{O}$. We say that $kv_0$ is an \textbf{even-odd} vector if $x,y$ are relatively prime and of opposite parity. We denote the \textbf{set of all even-odd directions} by $\mathcal{E}$.
\end{Def}
\noindent Let $\cS=\cO\sqcup\cE$ be the disjoint union of the two. This paper aims to depict the relationship between initial trajectory angles and dynamical properties of a geodesic on the surface. Recall that a geodesic on the surface is \emph{periodic} if there exists some $T>0$ such that $\Phi(t+T)=\Phi(t)$ for all $t\in\RR$ and a geodesic is \emph{drift-periodic} if there exists some $T'>0$ and non-trivial translation, $p$, such that $\Phi(t+T')=p(\Phi(t))$ for all $t\in\RR$. In either case we call this value the \emph{period} of $\Phi$ if it is the \emph{smallest} possible value that satisfies this definition. We additionally require that at no point does $\Phi$ encounter a cone singularity of the surface. Otherwise the geodesic behavior is undefined. We call $\Phi$ \emph{non-singular} if this holds.

\begin{align}
\text{Let }\XX'=\left\< \left[ \hspace{1mm} \begin{matrix}
							1 & 2\\
							 0 & 1
						\end{matrix}\hspace{1mm}\right],~
\left[ \hspace{1mm} \begin{matrix}
							1 & 0\\
							 2 & 1
						\end{matrix}\hspace{1mm}\right]\right\>
\end{align}

\noindent Section [blank] explains the significance of this group. Let $M\in\XX'$.

\begin{thm*}
(Dynamical Classification of Geodesics on the Necker Cube Surface) \\Let $\Phi_t:\RR\rightarrow\bUs$ be a non-singular unit-speed geodesic on the Necker cube surface such that $\Phi(0)=u_0$ and $\Phi'(0)=v\in\RR^3$. Let $v_0\RR^2$ be a distance-preserving projection of $v$ onto the plane tangent to the the surface at $u_0$. Let $\overline{u}_t\in\RR^3$ be $\Phi_t$'s position at time $t$. Let $k\in\RR$ such that $kv_0\in\cS$. Then the following is true:
\begin{enumerate}[label=(\roman*)]
\item $\Phi$ is periodic if $kv_0\in\cO$.
\item Suppose $\Phi$ is periodic. If $M (kv_0)=(1,1)$ then $\Phi$ has period $T=6\sqrt{2}||kv_0||$.
\item $\Phi$ is drift-periodic if $v_0\in\cE$. 
\item Suppose $\Phi$ is drift-periodic. If $M (kv_0)=(1,0)$ or $M (kv_0)=(0,1)$ then $\Phi$ has period $T=2||kv_0||$, and $d_E(\overline{u}_t,\overline{u}_{t+T})=||kv_0||\sqrt{2}$.
\end{enumerate}
\end{thm*}
\compav{Remember to prove that such an M exists.}

Here $||~||$ is the standard Euclidean norm in $\RR^2$, and $d_E:\RR^3\times\RR^3\rightarrow\RR$ is the Euclidean distance. The proof for this theorem can be found in XXX and follows from Theorems YYYY and ZZZZZ.

\subsection{Acknowledgements}
-Pat Hooper\\
- Vincent Delecroix, Ferrán Valdez, pascal hubert\\
- \\

\newpage

\section{Periodic Tiling of Necker Cubes}
This section will detail how the Necker Cube surface is constructed as a covering space of a thrice-punctured torus. Make the following identifications on 3 unit squares:
\begin{figure}[H]
\centering
\includesvg[width=1.83in]{neckercube}\includesvg[width=2in.]{neckercubepath}
\caption{$\bGs$ with and without paths $a,b,c,d$.}
\end{figure}

We will call this ``half-cube" $\bG$. It is a piecewise smooth surface homeomorphic to the two-torus, $\TT^2$. Every vertex has a cone angle of either $3\pi$ or $\frac{3\pi}{2}$. Let $\Sigma_{\bG}\subset\bG$ be the set of these three vertices. These are conical singularities of the surface. We denote the surface punctured at these points by $\bG\backslash\Sigma_{\bG} = \bGs$. The paths labeled $a,b,c,d$ are independent of one another and span $\bGs$'s fundamental group, $\pi_1(\bGs)\cong \<a,b,c,d \>$, the free group of four generators. The unpunctured Necker cube surface $\bU$ is embedded in $\RR^3$ by tiling along $\bG$'s outer edges:

\begin{figure}[H]
\centering
\includesvg[width=4in]{neckercubepathsurface}
\end{figure}
From this diagram we see that $\bU$ is a universal cover of $\bG$. In this paper we are studying geodesic behavior on $\bU$ and in order to do so, it is necessary to determine how the parallel transport of a unit vector around a cone singularity acts on the unit tangent bundle of the surface. But to study a given path's rotational holonomy, we require that these loops (see $c,d$ in Fig 3) aren't contractible through the cube's sharp corners. To that effect we puncture the surface at these points.  Denote the countably infinite set of singularities of $\bU$ by $\Sigma_{\bU}$. The \emph{Necker cube surface} is the punctured surface $\bUs=\bU\backslash\Sigma_{\bU}$. Observe from the previous diagram that $\pi_1(\bUs)$ is the kernel of the following group homomorphism:

\begin{Def}
Let $\varphi_1:\pi_1(\bGs)\rightarrow \ZZ^2$ be the group homomorphism given as $c,d\mapsto(0,0)$, $a\mapsto(1,0)$, and $b\mapsto(0,1)$.
\end{Def}

The cover of $\bGs$ with fundamental group $\ker\varphi_1$ is the space $\bUs=\tilde{\bGs}/\ker\varphi_1$ where $\ker\varphi_1$ acts on $\bGs$'s universal cover, $\tilde{\bGs}$, by deck transformations. We say $\Delta_{\varphi_1}=\pi_1(\bGs)/\pi_1(\bUs)\cong\ZZ^2$ is the \emph{deck group} of the covering map $\varphi_1^*:\bUs\rightarrow\bGs$. Now $\bGs$ is unfolded and flattened to better facilitate the study of its rotational holonomy. Apply the following cutting and gluing operations to the surface: \\
\begin{figure}[H]
\centering
\includesvg[width=3in]{LTorus}
\end{figure}
\noindent This L-shaped torus is then taken apart and recovered as $2\times 2$ torus missing a unit square with the following identifications:\\
\begin{figure}[H]
\centering
\includesvg[width=2in]{BranchTorus}
\caption{$\bGs$ and images of paths after reconstruction. The outermost vertex is \emph{not} a puncture.}
\end{figure}

Consider a $2\times2$ sheet missing an open unit square and that square's vertices (punctures). Center this structure around $0\in\CC$ and recover $\bGs$ with the identifications made in Fig 3. Let $m,n\in\ZZ$. Similar identifications can be made on the following subset of $\CC$ to recover $\bUs$:

\begin{equation}
\label{eq:P}
\mathbf{P} = \mathbb{C}\text{ }\backslash\bigcup_{m,n \in {\mathbb Z}} \big{\{} u+vi:~|u-2m|<\frac{1}{2},~|v-2n|<\frac{1}{2},~u=2m\pm\frac{1}{2}~v=2n\pm\frac{1}{2} \big{\}}.
\end{equation}

\compav{Include diagram of P}

\noindent Let $2m+i2n\in\CC$ be the center of a removed square with $m,n\in\ZZ$. Now let $\zeta_{m,n}(t)=(2m+t)+i(2n-\frac{1}{2})$ and 
$\xi_{m,n}(t)=(2m+t)+i(2n+\frac{1}{2})$ be paths on $\mathbf{P}$ that run along the top and bottom edges of the squares, parameterized by $t\in (-\frac{1}{2},\frac{1}{2})$ and $(m,n)\in\ZZ^2$. $\bUs$ is recovered as a topological quotient $\mathbf{P}/\sim$, where $\sim$ is the minimal relation on $\mathbf{P}$:

\begin{equation}
\begin{split}
\zeta(t)\sim e^{i\frac{\pi}{2}}~\overline{\zeta(t)-(2m+i2n)}\\
\eta(t)\sim e^{i\frac{\pi}{2}}~\overline{\eta(t)-(2m+i2n)}
\label{eq:rel2}
\end{split}
\end{equation}
The translations, $\pm 2,\pm 2i$, on $\mathbf{P}$ are isometries that preserve these identifications and induce isometries of $\bUs$. We denote the isometry induced by $+2$ as $f_h$, for horizontal, and the isometry induced by $+2i$ as $f_v$, for vertical. The group generated by these horizontal and vertical translations form the \emph{group of Deck transformations}, $\text{Deck}(\varphi_1^*)$, isomorphic to $\ZZ^2$:

\begin{figure}[H]
\centering
\includesvg[width=2.3in]{tiling}
\end{figure}

One can see that the Let $x_0\in\bGs$ and fix a point in the fiber,  $u_0\in\varphi_1^*{-1}(x_0)\subset\bUs$. The action of $\ZZ^2$ on the fiber over $x_0$ is given by $(m,n)\cdot u_0=(f_h^m\circ f_v^n)(u_0)$.

\compav{Discuss:}
The distance between $u_0$ and $(f_h^m\circ f_v^n)(u_0)$ is measured in its ambient space and taken to be the magnitude of $2m+i2n\in\CC$.

\begin{lem}
Let $\alpha:[0,1]\rightarrow \bGs$ be a loop on the surface. Let $[\alpha]\in\pi_1(\bGs)$ be its homotopy class representative. Denote its lift to $\bUs$ based at $u_0$ by $\tilde{\alpha}$. Then $\tilde{\alpha}(1)=\varphi([\alpha])\cdot\tilde{\alpha(0)}$.
\end{lem}

\subsection{Monodromy Group Representation}
A \emph{rotational monodromy map} is a morphism from the first fundamental group of a surface to the \emph{monodromy group} acting on the unit tangent space via parallel transport of a vector along its path. Existence of a parallel transport map is a consequence of the metric connection on a surface with complex structure.
\compav{Discuss:}\\
The monodromy map is well defined when the \emph{holonomy group} is \emph{discrete}. When the \emph{relative holonomy} of contractible loops is trivial, the holonomy group is the set of attainable rotations of a unit vector upon returning to the base point after transporting it along a non-trivial loop. A \emph{flat surface} like $\bGs$ has trivial relative holonomy as its neighborhoods are isometric to the plane. Denote the \emph{unit tangent bundles} of $\bUs$ and $\bGs$ by $T^1\bUs$ and $T^1\bGs$. If $\alpha$ is a loop on $\bGs$, then $[\alpha]$ acts on a fixed vector $v_0\in\RR^2$ by rotation. In the case where $\alpha$ is \emph{not} a geodesic segment on the surface of length $L$, $v_0=\alpha'(0)\in\RR^2$ may differ from $\alpha'(L)$ by some rotation.

\begin{Def}
Define $\varphi_2:\pi_1(\bGs)\rightarrow SO(2,\ZZ)$ to be the group homomorphism where $a,b \mapsto I_2$, $c  \mapsto R$, and $d  \mapsto R^3$, where $I_2=\begin{bmatrix}1 & 0\\0 & 1\end{bmatrix}$ and $R=\begin{bmatrix}0 & -1 \\1 & 0\end{bmatrix}$.
\end{Def}

\begin{figure}[H]
\centering
\includegraphics[width=2.5in]{monodromy.png}
\caption{Effect a nontrivial loop has on basis vectors $(i,j)\in\RR^2$.}
\label{fig:loop}
\end{figure}

\begin{thm}
$\varphi_2:\pi_1(\bGs)\rightarrow \mathrm{SO}(2,\ZZ)$ is the rotational monodromy representation of $\bGs$.
\begin{proof}
From Figure 4 it's clear that $c,d$ will rotate a unit vector by $\pm 90^\circ$ by parallel transport along their paths. These values are determined by the cone angles of the singularities they loop around. The parallel transport of a vector along $a$ or $b$ would be trivial since the outer edges they pass through are glued by translation.
\end{proof}
\end{thm}

\compav{Discuss:}
\begin{lem}
Any element in $\text{Deck}(\varphi_1^*)$ acting on $T^1\bUs$ has a trivial effect on holonomy.
\begin{proof}
Since $\bGs=\bUs/\ZZ^2$, the deck group action lifts naturally to an action on their tangent vector spaces and unit tangent bundles via pushforward of deck transformations $f_h,f_v$ on the bundle maps. i.e. $T^1\bGs=T^1\bUs/\ZZ^2$. This effect must be trivial since $\varphi_1(a)\sim f_h$ and $\varphi_1(b)\sim f_v$, and both homotopy classes $a,b$ have trivial holonomy.  \\
How do $f_h,f_v$ parallel transport a vector?\\
Well if $\varphi_1([a])\cdot u_0=f_h(u_0)$, then for 
\end{proof}
\end{lem}

A ramified cover of $\bGs$ is obtained by cyclically gluing 4 copies of neighborhoods surrounding its cone singularities. This cover is ramified over three singularities of cone angle $\frac{3\pi}{2}$ and $3\pi$, so 4 is the minimal number of copies needed to give every singularity in the cover a cone angle that is a multiple of $2\pi$. We call this cover $\bMs$.

\begin{figure}[H]
\centering
\includegraphics[width=3in]{monogroup.png}
\label{fig:arbitrarylift}
\caption{Lift of paths $c,d$ to $\bMs$ (\compav{RELABEL EDGES})}
\end{figure}

\begin{Def}
Let $\bMs$ be the regular cover of $\bGs$ with fundamental group $\pi_1(\bMs)=\ker\varphi_2$.
\end{Def}
\compav{Mark the puncutres}\\
That $\bMs$ is a degree four cover of $\bGs$ with trivial rotational holonomy and deck group $\SO(2,\ZZ)$ follows immediately from its construction. The metric completion of $\bMs$ is the compact translation surface $\bM$ obtained by reinserting its cone singularities. A \emph{translation surface} is a topological surface with a discrete set of cone singularities and a maximal translation atlas on the surface with these singularities removed. Neighborhoods of the surface are naturally isometric to $\CC$, and a metric on the translation surface is obtained by pulling back the Euclidean metric in $\CC$.

\subsection{A Four-Fold Cover}
We now take these covers we built to construct a third one that covers them all. This is a four-fold cover of $\bUs$ and a $\ZZ^2$ translation cover of $\bMs$. 
\begin{Def}
Let $\btUs$ be the cover of $\bUs$ with fundamental group $\pi_1(\btUs)=\ker\varphi_2\big\vert_{\pi_1(\bUs)}$.
\end{Def}

It follows that $\ker\varphi_1\big\vert_{\pi_1(\bMs)}=\ker\varphi_2\big\vert_{\pi_1(\bUs)}$ and that $\btUs$ is a cover of these two surfaces. Edge identifications are made on four copies of $\mathbf{P}$, possibly indexed by the finite component of the direct product $\mathbf{P}\times\ZZ/4\ZZ$. The Deck group of the covering map from $\btUs$ to $\bGs$ is the set $\ZZ^2\times \SO(2,\ZZ)$. Consider a loop on $\bUs$:

\begin{figure}[H]
\centering
\includegraphics[width=2in]{overlay.png}
\end{figure}

\noindent If we lift this curve to $\btUs$ we can identify it with an element of the deck group, and check if its lift from $\bGs$ belongs to the kernel of both $\varphi_1$ and $\varphi_2$. Fix a direction on this surface pointing north that respects $\bUs$'s compass and rotate every plane by either $90^\circ,$ $180^\circ,$ or $270^\circ$ as the curve travels from one plane (copy of $\mathbf{P}$) to the next. Observe that it will not only close if the curve acts trivially on the holonomy of $\bUs$, but with this construction we can see that $\btUs$ is actually an infinite-type translation surface:

\begin{figure}[H]
\centering
\includegraphics[width=3in]{coverdirection.png}
\caption{With $P\in 2\ZZ+2\ZZ i$.}
\end{figure}

\noindent

\subsection{Geodesics on $\bUs$}
The key connection made between surfaces $\bUs$ and $\btUs$ is that geodesics on $\bUs$ and $\btUs$ behave identically. A \emph{geodesic} is a \emph{unit-speed} locally distance minimizing curve on a smooth surface as a continuous function parameterized by an element in $\RR$. A \emph{geodesic segment} is a restriction of the geodesic to some closed interval in $\RR$. A geodesic has derivative defined at time 0, represented by a unit vector $v_0\in\RR^2$. If the vector can be scaled by $k\in\RR$ such that $kv_0\in\cS$ (defined in the introduction), the geodesic has a \emph{rational} initial trajectory angle. For example, the following is a periodic geodesic on the surface with initial direction $kv_0\in\cO$:
\\\compav{find this direction}
\begin{figure}[H]
\centering
\includegraphics[width=4in.]{closed2.png}
\caption{Periodic geodesic path on $\bUs$ modeled with sage-flatsurf.}
\label{fig:complicated}
\end{figure}

We refer later to the following commuting diagram of branched covering maps and inclusion maps. It is meant to illustrate the relationship between all of these spaces.

\begin{equation}
\begin{tikzcd}
\arrow[two heads, bend left]{rrr}{f^\circ}[swap]{} \arrow[hook]{d}{pr_b} \btUs \arrow[two heads]{r}{r^\circ} & \bUs \arrow[hook]{d}{pr_c} \arrow[two heads]{r}{q^\circ} & \bGs \arrow[hook]{d}{pr_d} &\arrow[two heads]{l}{w^\circ} \bMs \arrow[hook]{d}{pr_a}  \\
\arrow[two heads, bend right]{rrr}{}[swap]{f} \btU \arrow[two heads]{r}{r} & \bU \arrow[two heads]{r}{q} & \bG & \arrow[two heads]{l}{w}\bM  \\
\end{tikzcd}
\label{eq:commute}
\end{equation}

Let $x_0\in G^\circ$ and fix $u_0\in\bUs$, an element in the fiber of the covering map. A \emph{geodesic} on $G^\circ$ is a closed unit-speed straight-line path given as $\eta:\RR\rightarrow\bGs$. The \emph{period}/\emph{length} of the geodesic is some strictly positive $b\in\RR$. Let $\eta|_{[0,b]}=\alpha:[0,b]\rightarrow\bGs$ be a closed geodesic segment on $\bGs$ with initial and terminal point $x_0$. Denote $\alpha$'s unique lift to $\bUs$ by $\gamma:[0,b]\rightarrow\bUs$, where $\gamma(0)=u_0$, and $\gamma(b)$ in the fiber over $x_0$. Fix a point $\utild\in\btUs$ in the fiber over $u_0$. We define the unique lift of $\gamma$ to be the curve $\tilde{\gamma}:[0,b]\rightarrow\btUs$ with $\tilde{\gamma}(0)=\utild$. Denote their homotopy class representatives by $[\alpha]$ and (if they exist) $[\gamma],~[\tilde{\gamma}]$.

The geodesic segment on $\bGs$ has the property that it returns to $x_0$ with a trivial effect on the holonomy of the curve. On the other hand, the lifted paths will close only if their homotopy classes belong in their fundamental groups, i.e. the kernel of homomorphisms $\varphi_1$, $\varphi_2|_{\ker\varphi_1}$. Otherwise they are mapped non-trivially to $\ZZ^2$ or $\ZZ^2\times\SO(2,\ZZ)$. Let $v_0\in\RR^2$ be a unit vector such that $\alpha's$ derivative is $v_0$ at times $0$ and $b$. The derivative of $\gamma$ is also $v_0$ at these times as per Corollary 1, as $\alpha$ is a geodesic with trivial holonomy. $\gamtild'$ are also $v_0$ at their endpoints.

The Deck group of the map $\btUs\rightarrow\bUs$ is $\SO(2,\ZZ)$. This deck group acts naturally on the tangent bundles via push-forward of rotational matrices. We obtain the identity $T^1\btUs/\SO(2,\ZZ)\cong T^1\bUs$. This action is \emph{non-trivial}. Essentially, this means that there are 4 choices of a vector on the four-fold cover given a vector $v_0$ on $\bUs$. The choice of vector does not matter in the case where we flow it along a geodesic path, as a geodesic must return to a point in the fiber of $u_0$ in the same direction. That is to say that no matter where we start by lifting this path, we will always end up on the same plane (copy of $\mathbf{P}$) that we started in. In the case where a geodesic is periodic on $\bUs$, it would return to the same point in the fiber of $u_0$ as well. For simplicity's sake we will just lift to the vector mod the identity matrix, and call it $v_0$ as well.

\begin{Def}
Let $\Phi_t:\RR\rightarrow\bUs$ be a unit-speed geodesic on the \emph{Necker cube surface} with $\Phi(0)=u_0$ and $\Phi'(0)=v_0$. \\Similarly, let $\tilde{\Phi}_t:\RR\rightarrow\btUs$ be a unit-speed geodesic on $\bUs$'s four-fold cover with $\phitild(0)=\utild$ and $\phitild'(0)=v_0$.
\end{Def}
Let $\varphi=\varphi_1\times\varphi_2:\pi_1(\bGs)\rightarrow \ZZ^2\times\SO(2,\ZZ)$ be given by $[g]\mapsto(\varphi_1([g]),\varphi_2([g]))$. $\varphi$ is a
product of group homomorphisms that takes a loop to 
\begin{thm}
Let $\delta=\varphi_1([\alpha])\in\ZZ^2$ be an element of $\Delta_{\varphi_1}$ and $\tilde{\delta}=\varphi([\alpha])\in\ZZ^2\times \SO(2,\ZZ)$ be an element of $\Delta_{\varphi}$. Then:\\
$\Phi(t+b)=\delta\cdot\Phi(t)$ for all $t\in\RR$, and 
$\phitild(t+b)=\tilde{\delta}\cdot\phitild(t)$ for all $t\in\RR$.
\end{thm}

\begin{cor}
The following are equivalent:
\begin{enumerate}
\item $\Phi$ is periodic.
\item $\phitild$ is periodic.
\item $\delta=(0,0)$.
\end{enumerate}
\end{cor}

\begin{cor}
The following are equivalent:
\begin{enumerate}
\item $\Phi$ is drift-periodic.
\item $\phitild$ is drift-periodic.
\item $\delta\neq(0,0)$.
\end{enumerate}
\end{cor}

These theorems essentially tell us that these types of geodesics behave identically on the Necker cube surface and its cover. Now we show that this is true for $\btUs$ and its metric completion, $\btU$.

>>>>

\documentclass[a4paper, 11pt]{article}
\usepackage{comment} % enables the us`e of multi-line comments (\ifx \fi) 
\usepackage{lipsum} %This package just generates Lorem Ipsum filler text. 
\usepackage{fullpage} % changes the margin
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{svg}
\usepackage{enumitem}
\usepackage{color}
\usepackage{float}
\usepackage[utf8]{inputenc}
\usepackage{wrapfig}
\usepackage[arrow]{xy}
\usepackage{tikz-cd}

@@

\title{On the Characterizations of Rational Geodesics on the Infinite Necker Cube Surface}
\author{Pavel Javornik}
\date{}
\maketitle

%\begin{center}
%
%\includesvg[width=4.8in]{cubecoverphoto}\\
%\end{center}
\begin{abstract}
\noindent The deterministic quality of the dynamical characteristics, up to periodicity and drift periodicity, of a non-singular, rational geodesic on the infinite Euclidean cone surface obtained from the spatial edge-to-edge arrangement of infinitely many unit cubes comes down to the choice of an initial trajectory angle as a $\ZZ^2$ vector contained in the plane tangent to the surface at the geodesics initial point. Moreover these integer components give up the total period of a unit-speed geodesic, and their parities ultimately dictate whether or not it will be periodic or drift periodic. Using the surface and its translation cover's various symmetries, we prove these results and show that, when applicable, the Euclidean drift distance per period is given up by an element in the solution space of a system of $\ZZ$-linear equations.
\end{abstract}
\section{Introduction}
The Necker cube has been studied extensively due to its interesting effects on human visual perception. A popular adaptation of the Necker cube is the edge-to-edge pairing of multiple solid cubes giving the illusion of a rhombile tiling of the plane when viewed from a fixed angle. (Fig \ref{fig:front}) It is most recognizable in the works of M.C. Escher \cite{Escher1}\cite{Escher2} and as the game board in the 80s cabinet arcade classic, $\text{Q}^*\text{bert}$.\cite{Qbert} The cube was named after the Swiss crystallographer Louis A. Necker for having been one of the first to study the wire-frame cube.\cite{Albert}
\begin{figure}[H]
\centering
\includesvg[width=5.5in]{oddoddtrajectory}
\caption{Periodic Geodesic on the Necker Cube Surface}
\label{fig:front}
\end{figure}
We take the three visible faces of a single cube viewed from this angle and make identifications on the parallel edges by translation. (Fig \ref{fig:halfcube}) These are the outer edges that are glued to the opposite edges of other cubes. We call this half-cube $\bG$. $\bG$ has a set of three cone singularities, $\Sigma_\bG$, two with cone angle $3\pi$ and one with cone angle $\frac{3\pi}{2}$. The half-cube punctured at these points is $\bGs=\bG\backslash\Sigma_\bG$, and we call this our \emph{Necker cube}.
\begin{figure}[H]
\centering
\includesvg[width=1.83in]{neckercube}\includesvg[width=2in.]{neckercubepath}
\caption{The surface $\bGs$ with and without loops $a,b,c,d$ based at $x_0$.}
\label{fig:halfcube}
\end{figure}
The set of paths labeled above generate the fundamental group of the surface, $\pi_1(\bGs,x_0)$, that is isomorphic to the free group of four generators since $\bG$ is homeomorphic to the genus 1 torus. The \emph{Necker cube surface} itself is a cover of $\bGs$ with a deck group isomorphic to $\ZZ^2$:

\begin{figure}[H]
\centering
\includesvg[width=4.3in]{neckercubepathsurface}
\end{figure}

\begin{Def}
Let $\varphi_1:\pi_1(\bGs)\rightarrow \ZZ^2$ be the homomorphism on the free group $\<a,b,c,d\>$ given by $c,d\mapsto(0,0)$, $a\mapsto(1,0)$, and $b\mapsto(0,1)$.
\end{Def}

The cover of $\bGs$ with fundamental group $\ker\varphi_1$ is the space $\bUs=\tilde{\mathbf{G}}^\circ/\ker\varphi_1$ where $\ker\varphi_1$ acts on $\bGs$'s universal cover,  $\widetilde{\mathbf{G}}^\circ$, by deck transformations. We can see that $\bGs$ is obtained as the quotient of the Necker cube surface by these $\ZZ^2$ symmetries.

\begin{Def}
Let $p_1:\bUs\rightarrow\bGs$ be a projection that induces a monomorphism on the fundamental groups of these surfaces such that $p_{1*}\pi_1(\bUs)=\ker\varphi_1$.
\end{Def}
Fix some point $u_0$ in $\bUs$ in the fiber over $x_0$, $p_1^{-1}(x_0)$. These are both contained in smooth sections, i.e. a face of a cube.
\begin{Def}
Define $T_{u_0}(\bUs)$ as the vector space tangent to the surface at $u_0$. We denote the unit tangent space at this point $T^1_{u_0}(\bUs)$. Both belong to their respective tangent bundles, $T(\bUs)$ and $T^1(\bUs)$.
\end{Def}

\subsection{Discussion of Results}
For every direction in the tangent space at $u_0$ there is a unique geodesic on $\bUs$.
\begin{Def}
Let $\Phi_t:\RR\rightarrow\bUs$ be the unit-speed geodesic with $\Phi(0)=u_0$ and $\frac{d}{dt}\Phi(0)=v_0\in T^1_{u_0}(\bUs)$. It is the image of the geodesic flow based at $u_0$ on the unit tangent bundle in direction $v_0$. We additionally require that at no point does the geodesic encounter a cone singularity of the surface.
\end{Def}
We say that $\Phi$ is \emph{periodic} if for some $T>0$, $Phi(t+T)=\Phi(t)$. Let $f$ be a non-trivial automorphism of the surface. We say that $\Phi$ is \emph{drift-periodic} if for some $T'>0$, $\Phi(t+T')=f(\Phi(t))$. The initial trajectory angle is said to be $\emph{rational}$ if there is some $k\in\RR$ such that $kv_0$'s non-zero components are relatively prime integers. Although $v_0$ is a vector of $\RR^3$, the tangent spaces are parallel to subspaces generated by exactly two standard basis vectors. For that reason, one component is always 0 and we will just treat it as an element of $\RR^2$. A rational initial trajectory angle in $\RR^2$ falls into one of two categories:
\begin{Def}
Let $v_0$ be a rational unit vector of the form $\frac{1}{k}(x,y)\in\RR^2$ with $x,y\in\mathbb{Z}$ and $k=\sqrt{x^2+y^2}\in\mathbb{R}$. We say $kv_0$ is an \textbf{odd-odd} vector if $x,y$ are relatively prime and both odd. We denote the \textbf{set of all odd-odd directions} $\mathcal{O}$. We say that $kv_0$ is an \textbf{even-odd} vector if $x,y$ are relatively prime and of opposite parity. We denote the \textbf{set of all even-odd directions} $\mathcal{E}$. Denote $\cS=\cO\sqcup\cE$ as the \textbf{collection of all possible rational directions}.
\end{Def}

We use $||~.~||:\RR^2\rightarrow\RR_{\geq 0}$ to denote the standard Euclidean norm in $\RR^2$.

\begin{thm*} \textbf{Periodic and Drift-Periodic Geodesics on the Necker cube surface.} 
\\Let $\Phi_t:\RR\rightarrow\bUs$ be a non-singular unit-speed geodesic on the Necker cube surface such that $\Phi(0)=u_0$ and $\Phi'(0)=v\in\RR^3$. Let $v_0\in\RR^2$ be a distance-preserving projection of $v$ onto the plane tangent to the surface at $u_0$. Let $k\in\RR$ such that $kv_0\in\cS$. Then the following is true:
\begin{enumerate}[label=(\roman*)]
\item $\Phi$ is periodic if $kv_0\in\cO$ with period $T=6||kv_0||$.
\item $\Phi$ is drift-periodic if $v_0\in\cE$ with period $T=2||kv_0||$.
\end{enumerate}
\end{thm*}
\newpage
\newpage
\begin{thebibliography}{9}
\bibitem{Necker} 
\bibitem{Albert} Necker, L.A., \emph{Observations on some remarkable optical phaenomena seen in Switzerland; and on an optical phaenomenon which occurs on viewing a figure of a crystal or geometrical solid.} London and Edinburgh Philosophical Magazine and Journal of Science. 1832. doi:10.1080/14786443208647909
\bibitem{Qbert} Gottlieb., \emph{Q*bert} [Arcade], Gottlieb, 1982.
\bibitem{Escher1} Escher, M.C., \emph{Metamorphosis I}, 1937.
\bibitem{Escher2} Escher, M.C., \emph{Convex and Concave}, 1955.
\bibitem{flatsurf} include github link
\end{thebibliography}
\end{document}
